Data Provenance for Concurrent Program Debugging and Analysis in Production Systems

### 2013-11-12 paper.tex
committing an intermediate state on the ride home...

Prior work have previously shown that some limited forms of provenence help
with debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort
tracked code that wrote unusable values into variables~\cite{badapples}.  That
work showed that revealing the connection between a bad value's origin and a
failure it causes is helpful during debugging.  Other work showed that
programmers debug programs more effectively when they can explicitly ask "why"
questions about the origin of certain values during
debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea.  First, prior work~\cite{badapples} has focused on providing
provenance information for only some unusable {\em values}, like {\tt NULL}.
Limiting the applicability of their technique to only some failures, like {\tt
Null Pointer Exceptions}.  Second, while prior work has shown that limited
provenance tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples} impose overheads too high for production for C
and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Addressing
concurrency is important because a value could have provenance in any of a
large number of threads, making reasoning about why a variable contains that
value complex and difficult.

@@

Last writer slices are primarily useful for concurrency debugging, but we also
observe that provenance information in the form of last writer slices is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work has used {\em ad hoc} last-writer information as part of
more heavy-weight analyses in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conmem,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
Traps} or {\em \ctraps}  on top of last writer slices
(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each
potentially shared memory location when a thread accesses such a location.  If
the accessing thread is different from the thread that last wrote the location,
the access signifies {\em communication} between the threads.  When
communication occurs, \ctraps delivers a software {\em trap} to the accessing
thread giving it information about how communication occurred.  The trap can be
handled by an analysis-specific trap handler.  Trap handlers implement {\em
\ctraps applications} through an extensible interface that can interpose on the
execution (Section~\ref{sec:apps}).  By default, \ctraps has very low overheads
and similarly to last writer slicing are appropriate for use in production in
many cases.  \ctraps applications can perform arbitrary concurrency analyses,
but at their own discretion and cost.  As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.

@@

\section{Last Writer Slices}
\label{sec:lastwriterslices}
A last writer slice is a dynamic property of a memory location at a point in a
program's execution.   A memory location's last writer slice is a record
containing the program point and the identifier of the thread that last wrote
data to that memory location.  We gather last writer slices at runtime by
maintaining a {\em last writer table} (\lwt) that maps from a memory
location's address to its last writer slice.  Just before a thread writes to a
memory location,  the thread first finds the memory location's entry in the
\lwt and updates that entry with its thread identifier and its current point in the
program.  After updating the \lwt, the thread performs the original memory
access.

@@

\subsection{Runtime \& Compiler Support}
 The \lwt is implemented as a {\em shadow memory} that holds one entry for
each memory address accessed by the program.  Each \lwt entry records the
thread and program point that last wrote to the entry's corresponding memory
location.  When a thread performs a write to a memory location, it updates the
\lwt by recording its thread identifier and the program point of its write
operation to the memory location's \lwt entry.  A program point may be an
instruction address, or a call stack.  We use instructions by default, but we experimented with 2-address bounded call-stacks as well.

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call into our runtime
system just before each write to a memory location that may be accessed by code
executing in another thread (escaping memory accesses).  Write instrumentation
calls are passed three pieces of information: the address of the memory word
involved in the access, the thread identifier of the accessing thread, and the
accessing program point.

\subsubsection{Escape Analysis}
We use a local function escape analysis to identify which accesses are to
memory locations that can be statically guaranteed to be inaccessible to code
running in other threads.  When a thread accesses a location that is
inaccessible to code in other threads, there is no way that access can be a
communicating access.  Such accesses are not instrumented by our compiler.  
We rely on {\tt GCC}'s built-in function escape analysis to determine if a
memory operation might lead to communication.  Locations determined not to
escape function scope also cannot escape the thread executing the function.
%Using an inter-procedural thread escape analysis may eliminate more
%instrumentation sites.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=.90\columnwidth]{figs/SystemDiagram.pdf}
\caption{\label{fig:systemdiagram}CTraps system support.}
\end{figure}

>>>>

Prior work has previously shown that some limited forms of provenence help with
debugging~\cite{badapples, whylineicse, tipslicingsurvey}.  One effort tracked
code that wrote unusable values into variables~\cite{badapples}.  That work
showed that revealing the connection between an unusable value's origin and a
failure it causes is helpful during debugging.  Other work showed that
programmers debug more effectively when they can ask "why" questions about
values during debugging~\cite{tipslicingsurvey,whylineicse}.

These efforts have shown some of the promise of provenance, but are not a
debugging panacea.  First, prior work~\cite{badapples} has focused on providing
provenance information for only some unusable {\em values}, like {\tt NULL}.
Limiting the applicability of their technique to only some failures, like {\tt
Null Pointer Exceptions}.  Second, while prior work has shown that limited
provenance tracking for Java programs is possible with overheads low enough for
production~\cite{badapples}, a more general form of provenance tracking for C
and C++ programs has been elusive.  Prior approaches that use
slicing~\cite{thinslicing, tipslicingsurvey, whylineicse} or heavy-weight
memory checkers~\cite{badapples} impose overheads too high for production for C
and C++ programs. Third, no prior technique that we know of has collected
provenance information that is tailored to debugging concurrent programs with
overheads that are low enough for use in production systems.  Addressing
concurrency is important because a value could have originated in any of a
large number of threads, making reasoning about why a variable contains that
value complex and difficult.

@@

Last writer slices are primarily useful for concurrency debugging, but we also
observe that provenance information in the form of last writer slices is a
useful foundation for implementing other concurrent program analyses.  Several
tools from prior work has used {\em ad hoc} last-writer information as part of
more heavy-weight analyses in domains like bug detection ({\em e.g.},
~\cite{dmtracker,avio}), automatic debugging ({\em e.g.},
~\cite{recon,conseq,cci}), profiling ({\em
e.g.},~\cite{threadclustering, schedpredictionmodel}), and program
understanding~\cite{oshatr}.

Motivated by these prior {\em ad hoc} efforts, we build {\em Communication
Traps} or {\em \ctraps}  on top of last writer slices
(Section~\ref{sec:ctraps}).  \ctraps checks the last writer slice of each
potentially shared memory location when a thread accesses such a location.  If
the accessing thread is different from the thread that last wrote the location,
the access signifies {\em communication} between the threads.  When
communication occurs, \ctraps delivers a software {\em trap} to the accessing
thread, giving it information about how communication occurred.  The trap can be
handled by an analysis-specific trap handler.  Trap handlers implement {\em
\ctraps applications} through an extensible interface that can interpose on the
execution (Section~\ref{sec:apps}).  By default, \ctraps has very low overheads
and similarly to last writer slicing are appropriate for use in production in
many cases.  \ctraps applications can perform arbitrary concurrency analyses,
but at their own discretion and cost.  As we show in our evaluation, the
overhead of \ctraps applications scales with the complexity of their analysis.

@@

\section{Last Writer Slices}
\label{sec:lastwriterslices}
A last writer slice is a dynamic property of a memory location at a point in a
program's execution.   A memory location's last writer slice is a record
containing the program point and the identifier of the thread that last wrote
data to that memory location.

We gather last writer slices at runtime by maintaining a {\em last writer
table} (\lwt) that maps from a memory location's address to its last writer
slice.  Just before a thread writes to a memory location, the thread first
finds the memory location's entry in the \lwt and updates that entry with its
thread identifier and its current point in the program.  After updating the
\lwt, the thread performs the original memory access.

@@

\subsection{Runtime \& Compiler Support}

Figure~\ref{fig:lwsoverview} shows an overview of our system support for last
writer slicing.  The programmer writes their program as usual. Our compiler
instruments write operations in the program with calls to our runtime that
update the \lwt.  The \lwt is maintained in the runtime and saved for use
during debugging, like the core dump.

\begin{figure}[h]
\centering
\includegraphics[width=.9\columnwidth]{figs/LWS_Overview.pdf}
\caption{\label{fig:lwsoverview}{\bf Overview of system support for last writer slicing.} Our compiler and runtime library ensure the \lwt remains updated so it can be used during debugging.}
\end{figure}

The \lwt implemented in the runtime library is a {\em shadow memory} that
holds one entry for each potentially shared memory address accessed by the
program.  Each \lwt entry records the thread and program point that last wrote
to the entry's corresponding memory location.  When a thread performs a write
to a memory location, it updates the \lwt by recording its thread identifier
and the program point of its write operation to the memory location's \lwt
entry.  A program point may be an instruction address, or a call stack.  We use
instructions by default, but we experimented with 2-address bounded call-stacks
as well.

Collecting last writer slices requires instrumentation on each write operation.
We built an instrumenting compiler pass that inserts a call to an \lwt update
function in our runtime library just before each write that may access data
last written by another thread (escaping memory accesses).  Write
instrumentation calls are passed three pieces of information: the address of
the memory word involved in the access, the thread identifier of the accessing
thread, and the accessing program point.

We use a {\tt GCC}'s local function escape analysis to identify accesses to
memory locations that are provably inaccessible to code running in other
threads.  Such operations access unshared data and they are not instrumented by
our compiler.  
%Using an inter-procedural thread escape analysis may eliminate more
%instrumentation sites.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=.90\columnwidth]{figs/CTraps_Overview.pdf}
\caption{\label{fig:systemdiagram}{\bf CTraps system support.}}
\end{figure}
